---
title: 'Final Project: The Socioeconomic and Food Reatil Landscape of San Francisco''s
  Mission District'
authors: Fischer Heimburger (@fheimburger97) & Caroline Mahdavi (@cmahdavi)
output:
  github_document:
    df_print: tibble
  pdf_document: default
---
By: Fischer Heimburger (@fheimburger97) & Caroline Mahdavi (@cmahdavi)

# Introduction
The Mission District (known as "This Mission") in San Francisco known for it's widening economic inequity, relatively high ethnic diversity--including being a local hub for Latinx culture and heritage, and vibrant food scene. This research seeks to first cultivate an understanding of the socio-economic landscape of the Mission District, and then initiate research on how this landscape is reflected in the distribution of food retailers.

Economic inequity is a major challenge facing the Mission District today. In 2020, it was found that "the highest and lowest income brackets compose almost 40% of all Mission households" ([Mission Action Plan 2020](https://default.sfplanning.org/Citywide/Mission2020/MAP2020_Status_Report_2018.pdf). This demonstrates the increasing income gap experienced in the Mission District. In addition, the economic landscape of the Mission District is highly influenced by gentrification--the process by which the migration of high-income residents to a place displaces low-income, often marginalized, residents ([Chapple and Thomas 2021](https://www.urbandisplacement.org/about/what-are-gentrification-and-displacement/)). San Francisco was recently found to be the most intensely gentrified city in America ([Richardson, Mitchell, and Edlebi 2020](https://ncrc.org/gentrification20/)) and the Mission District, one of the neighborhoods with the most advanced levels of gentrification ([Chapple and Thomas 2021](https://www.urbandisplacement.org/about/what-are-gentrification-and-displacement/)). We hope to understand this unique and stark economic landscape by mapping out median household incomes by census tract within the Mission District. Our research on this is influenced by data and figures from the [Urban Displacement Project](https://www.urbandisplacement.org/). The Urban Displacement Project traces gentrification throughout the San Francisco Bay Area. As a part of their studies, they gather data on median income levels within the Mission District and map out their findings. We hope, in part, to recreate aspects of their figures with updated census data.

While much of the Bay Area is growing more racially segregated ([Menendian, Gambhir and Gailes 2021](https://belonging.berkeley.edu/roots-structural-racism)), The Mission is unique for its relatively high ethnic diversity ([Hom 2021](https://missionlocal.org/2021/07/the-mission-isnt-as-racially-segregated-as-the-bay-does-that-point-to-gentrification)). While overall in 2019 the Mission District was 38.7% Latinx, 36.4% White, and 13.7% Asian ([City Data 2021](https://www.city-data.com/neighborhood/Mission-District-San-Francisco-CA.html)); the demographic makeup of race varies depending on locale ([Hom 2021](https://missionlocal.org/2021/07/the-mission-isnt-as-racially-segregated-as-the-bay-does-that-point-to-gentrification)). Additionally, recent years have shown dynamic shifts in demographics as the 2020 Census revealed a 14.4% decline in Latinx residents within the Mission District and a 34.8% increase in Asian residents ([Horowitz and Jarret 2021](https://missionlocal.org/2021/09/census-2020-as-san-francisco-grew-the-ethnic-makeup-of-its-neighborhoods-changed-heres-how)). The [Othering & Belonging Institute](https://belonging.gis-cdn.net/us_segregation_map/?year=2020) researches levels of segregation in the San Francisco Bay Area. As part of their research, they have created maps of the Mission District that display levels of segregation and racial diversity. We seek to create similar maps that display racial majorities by census tract to further understand the demographic landscape of The Mission. We then hope to understand how these socio-economic landscapes interact the landscape of food retailers in the Mission District.

Food systems are complex, interconnected bio-physical and socio-economic webs that are influenced by environmental, social, political, and economic systems, institutions, and actors ([Ericksen 2008](https://doi.org/10.1016/j.gloenvcha.2007.09.002)). For many living in cities, like those within the Mission District, relationships with food systems revolve around food retailers—including supermarkets, grocery stores, corner markets, and convenience stores—that help connect food producers with food consumers ([Trivette 2019](https://link.springer.com/article/10.1007/s10460-018-9885-1)). Using data from the [City of San Francisco](https://datasf.org/opendata/), we hope to map out the current distribution of food retailers in the Mission District. Then we seek to understand how this distribution may be interconnected with class and race.

Our research is organized as follows. First, we map out income levels by census tract in the Mission District. Second, we map out racial majorities by census tract in The Mission. We then create a demographic table that organizes the socio-economic landscape of the district. In the fourth section we map out the distribution of food retailers within the Mission District. Lastly, we overlay the socio-economic maps with the food retailer map.

#### Data note: We used the following libraries to analyze our data. Additionally, please run `install.packages("tidycensus")` in the console window.
```{r setup, include=FALSE}
library(rgdal)
library(RSQLite)
library(tmap)        
library(terra)
library(tidyverse)
library(sf)         
library(ggplot2)
library(dplyr)
library(knitr)
library(reshape2)
library(httr)
library(jsonlite)
library(data.table)
install.packages("tidycensus")
install.packages("ggmap")
library(tidycensus)
library(ggmap)

knitr::opts_chunk$set(echo = TRUE)
```


Google API Key -- need to upload key
#```{r}
register_google(key = "", write = TRUE)
#```

## I Tract Income Typology

#### Table 1.1: Downloading San Francisco's Geopackage
First, we downloaded the [geopackage for San Francisco](https://github.com/urban-displacement/displacement-typologies/tree/main/data/downloads_for_public). In the following chunk, we load the geopackage using `read_sf` and `filter` for census tracts in the Mission District. We `select` to show "GEOID" and "geom" which will help create our map.
```{r}
gfile <- read_sf("sanfrancisco.gpkg") %>%
  select(GEOID, geom) %>%
  filter(GEOID == "6075020200" | GEOID == "6075020100" | GEOID == "6075020800" | GEOID == "6075020700" | GEOID == "6075022902" | GEOID == "6075020900" | GEOID == "6075017700" | GEOID == "6075022901" | GEOID == "6075022803" | GEOID == "6075022903" | GEOID == "6075022801" | GEOID == "6075021000" | GEOID == "6075022802")
```

#### Table 1.2: Downloading San Francisco's Demographics
We then use `read.csv` to load typology data from the [Urban Displacement Project github](https://github.com/urban-displacement/displacement-typologies/tree/main/data/outputs/typologies). The definitions for variable names can be found [here](https://github.com/urban-displacement/displacement-typologies/blob/main/data/outputs/typologies/typologies_codebook.md).

```{r}
demographics <- read.csv("https://raw.githubusercontent.com/urban-displacement/displacement-typologies/main/data/outputs/typologies/SanFrancisco_typology_output.csv")
```

#### Table 1.3: Median Household Income Table
First, we create the **medianincome** table which shows "GEOID" and "hinc_18"; "hinc_18" is the median household income in the past 12 months (in 2018 inflation-adjusted dollars) in 2018. Second, with `left_join`, we join the newly created **medianincome** table and the initial **gfile** table to create our **incomejoin** table.
```{r}
medianincome <- demographics %>%
  select(GEOID, hinc_18) %>%
  rename("Income in Dollars" = hinc_18)
incomejoin <-left_join(gfile, medianincome, by = "GEOID")
incomejoin
```

#### Figure 1.1: Mapping Median Household Income in the Mission District
Using **tmap** and our **incomejoin** table, we then create a map showing median household income across census tracts in the Mission District.
```{r message=FALSE, warning=FALSE}
incomemap <- tm_shape(incomejoin) +
             tm_style("watercolor") +
             tm_polygons("Income in Dollars") +
             tm_layout(main.title="Median Household Income",
                       main.title.position = "centre",
                       main.title.size = 1.6) +
             tm_legend(position = c("right", "top"),
             legend.outside = TRUE,
             legend.outside.size = .35,
             legend.title.size = 1.5,
             legend.text.size = 1.2) +
  tm_compass(position = c("left", "top"))
incomemap
```

## II Tract Racial Typology

To work with census data, we install **tidycensus** from CRAN with the following command: `install.packages("tidycensus")`. After loading the **tidycensus** and **tidyverse** libraries, obtain a [Census API key](http://api.census.gov/data/key_signup.html). Note: 2020 decennial Census data use differential privacy, a technique that introduces errors into data to preserve respondent confidentiality, small counts should be interpreted with caution. 
```{r message=FALSE}
census_api_key("77bb8e57772d9321db5adcd03b9bf2c3bce563c3", install = TRUE, overwrite = TRUE)
```

#### Table 2.1 Population Numbers by Race

Using **tidycensus**, we will get the values for four variables: total population, White population, Asian population, and Hispanic or Latino population. The argument to the `geography` parameter is "tract", and by by specifying `state`, `county`, and `year` we get the desired population numbers for each census tract in San Francisco from 2020. Note that `geometry` is set to true in the first table to help create maps later.

```{r message=FALSE, warning=FALSE}
#White Population
white <- get_decennial(geography = "tract",
                      state = "CA",
                      county = "San Francisco",
                      year = 2020,
                      variables = c(white = "P1_003N"))
#Asian Population
asian <- get_decennial(geography = "tract",
                      state = "CA",
                      county = "San Francisco",
                      year = 2020,
                      variables = c(asian = "P1_006N"))
#Hispanic or Latino Population
latinx <- get_decennial(geography = "tract",
                      state = "CA",
                      county = "San Francisco",
                      year = 2020,
                      variables = c(latinx = "P2_002N"))
#Total Population
totalpop <- get_decennial(geography = "tract",
                      state = "CA",
                      county = "San Francisco",
                      year = 2020,
                      variables = c(totalpop = "P1_002N"),
                      geometry = TRUE)
```

#### Table 2.2: Race Percents

Next, with a series of `left_join`, we join the four previous tables by "GEOID". We `rename` and `select` several important variables for neatness and clarity. We use `mutate` to add three columns displaying the percentage of the total population that each race is.
```{r}
totalandwhite <- left_join(totalpop, white, by = "GEOID") %>%
    rename(tract = NAME.x,
           totalvar = variable.x,
           totalval = value.x,
           whitevar = variable.y,
           whiteval = value.y) %>%
  select(GEOID, tract, totalvar, totalval, whitevar, whiteval)
asianandlatinx <- left_join(asian, latinx, by = "GEOID") %>%
    rename(tract = NAME.x,
           asianvar = variable.x,
           asianval = value.x,
           latinxvar = variable.y,
           latinxval = value.y) %>%
  select(GEOID, asianvar, asianval, latinxvar, latinxval)
racetable <- left_join(totalandwhite, asianandlatinx, by = "GEOID") %>%
  mutate(percwhite = (whiteval/totalval)*100) %>%
  mutate(percasian = (asianval/totalval)*100) %>%
  mutate(perclatinx = (latinxval/totalval)*100)
racetable
```

#### Table 2.3: Race Percents and Majorities

Next, we `mutate` to add a "majority" column. In this column, we use `case_when` to show which of the three races has the highest percent of the population, and thus, is the majority of its tract.
```{r}
percents <- racetable  %>% 
  mutate(Majority = case_when(
    perclatinx > percasian & perclatinx > percwhite ~ "Hispanic or Latinx",
    percwhite > percasian & percwhite > perclatinx ~ "White",
    percasian > percwhite & percasian > perclatinx ~ "Asian")) %>%
  select(GEOID, tract, percwhite, percasian, perclatinx, Majority, geometry)
```

#### Table 2.4: Specifying the Boundaries

To narrow our scope, we then `filter` to only include census tracts in the Mission District of San Francisco.
```{r}
mission_geo_zero <- percents %>% pull(GEOID)
mission_percents_clean <- percents %>%
  mutate(GEOID = substring(mission_geo_zero, 2)) %>%
  filter(GEOID == "6075020201" | GEOID == "6075020202" | GEOID == "6075020101" | GEOID == "6075020102" | GEOID == "6075020801" | GEOID == "6075020802" | GEOID == "6075020701" | GEOID == "6075020702" | GEOID == "6075022902" | GEOID == "6075020900" | GEOID == "6075017700" | GEOID == "6075022901" | GEOID == "6075022803" | GEOID == "6075022903" | GEOID == "6075022801" | GEOID == "6075021000" | GEOID == "6075022802")
mission_percents_clean
```

#### Figure 2.1: Mapping Racial Majorities in the Mission District

-
```{r}
racemap <-
  tm_shape(mission_percents_clean) +
  tm_polygons("Majority") +
  tm_style("watercolor") +
  tm_layout(main.title = "Racial Typography",
            main.title.position = "centre",
            main.title.size = 1.6) +
  tm_legend(position = c("right", "top"),
            legend.outside = TRUE,
            legend.outside.size = .35,
            legend.title.size = 1.5,
            legend.text.size = 1.2) +
  tm_compass(position = c("left", "top"))
racemap
```

## III Demographics Table

#### Table 3.1
```{r}
incomejoin_chr <- incomejoin %>%
  mutate_at("GEOID", as.character)
incomejoin_chr
```

#### Table 3.2: Demographics Table
```{r}
finaljoin <- right_join(st_drop_geometry(incomejoin_chr), st_drop_geometry(mission_percents_clean), by = "GEOID") %>% 
  select(GEOID, tract, "Income in Dollars", Majority, percwhite, perclatinx, percasian) %>% 
  rename("Median Income ($)" = "Income in Dollars", Tract = tract, "White Percentage" = percwhite, "Hispanic/Latinx Percentage" = perclatinx, "Asian Percentage" = percasian, "Racial Majority" = Majority)
finaljoin
```

## IV Food Retailers

#### Table 4.1: Uploading business data from data.sfgov.org.
```{r}
SF_businesses <- read_csv("https://data.sfgov.org/api/views/g8m3-pdis/rows.csv?accessType=DOWNLOAD")
SF_businesses
```

#### Table 4.2: Business in the Mission with LIC Code H in it. LIC codes with H's are retail businesses, this was a crucial filtering step to finding food retailers.
```{r}
Mission_H_Codes <- SF_businesses %>%
   filter(str_detect(`LIC Code`, "H"), `Neighborhoods - Analysis Boundaries` == "Mission") %>%
  select(`Ownership Name`, `DBA Name`, `Street Address`, `Business Start Date`, `Business End Date`, `Mail State`, `NAICS Code`, `NAICS Code Description`, `LIC Code`, `LIC Code Description`, `Neighborhoods - Analysis Boundaries`)
Mission_H_Codes %>% select(`Ownership Name`, `DBA Name`, `LIC Code`, `LIC Code Description`)
```

#### List 4.1: The following is a list of LIC codes that are food retailers. This list includes grocery stores, corner markets, convienence stores, supermarkets, bakeries, and drug stores. This list does not include restaurants.
```{r}
H_strings <- c("H01", "H02", "H03", "H04", "H05", "H06", "H07", "H08", "H09", "H10", "H11", "H12", "H13", "H14", "H16", "H19", "H80", "H81", "H83", "H88", "H89", "H92")
```

#### Table 4.3: The follow is a table that includes data on all food retailers within the Mission District.
```{r}
Final_Mission_Food_Retailers <- SF_businesses %>%
  filter(`Neighborhoods - Analysis Boundaries` == "Mission") %>%
  filter(str_detect(`LIC Code`, paste(H_strings, collapse = "|")))
Final_Mission_Food_Retailers
```

```{r}
Clean_Final_Retail <- Final_Mission_Food_Retailers %>%
  select("Ownership Name", "DBA Name", "Street Address", "LIC Code Description")
Clean_Final_Retail
```
```{r}
Clean_Order_r <- Clean_Final_Retail[order(Clean_Final_Retail$"DBA Name"),]
Clean_Order_r
```
```{r}
write.csv(Clean_Order_r, "C:\\Users\\fheim\\Desktop\\Test\\People.csv", row.names = FALSE)
```



#### List 4.2: The following are the coordinates for each food retail store in the Mission District.
```{r}
coordinates <- Final_Mission_Food_Retailers %>% pull("Business Location")
```

#### Figure 4.1: The follow is a map of all food retailers within the Mission District. Each dot represents a food retailer. While this map does not include streets or the outline of the Mission Distict, maps in the following section will include an outline of the Mission District along with demographic information.
*
```{r}
geo_final <- Final_Mission_Food_Retailers %>%
  rename(geometry = `Business Location`) %>%
  mutate(geometry = as.list(geometry)) %>%
  st_as_sf(crs = "WGS84")
tm_shape(geo_final) + tm_dots()
```

## V Final Maps

#### Figure 5.1: Food Retailers and Median Household Income in the Mission District
```{r warning=FALSE}
dots <- tm_shape(geo_final) + tm_dots(col="black") + tm_add_legend(shape = "tm_dots", labels = ('Food Retailers'), col = "black")
incomemap_with_food_retailers <- tm_shape(incomejoin) +
             tm_style("watercolor") +
             tm_polygons("Income in Dollars") +
             tm_layout(main.title="Median Household Income & Food Retailers",
                       main.title.position = "centre",
                       main.title.size = 1.6) +
             tm_legend(position = c("right", "top"),
             legend.outside = TRUE,
             legend.outside.size = .35,
             legend.title.size = 1.5,
             legend.text.size = 1.2) +
  tm_compass(position = c("left", "top"))
incomemap_with_food_retailers + dots
```

#### Figure 5.2: Food Retailers and Racial Majorities in the Mission District
*
```{r warning=FALSE}
dots <- tm_shape(geo_final) + tm_dots(col="black") + tm_add_legend(shape = "tm_dots", labels = ('Food Retailers'), col = "black")
racemap_with_food_retailers <-
  tm_shape(mission_percents_clean) +
  tm_polygons("Majority") +
  tm_style("watercolor") +
  tm_layout(main.title = "Racial Typography & Food Retailers",
            main.title.position = "centre",
            main.title.size = 1.6) +
  tm_legend(position = c("right", "top"),
            legend.outside = TRUE,
            legend.outside.size = .35,
            legend.title.size = 1.5,
            legend.text.size = 1.2) +
  tm_compass(position = c("left", "top"))
racemap_with_food_retailers + dots
```

# Conclusion
We were able to create maps of the Mission District based off median household income, racial majorities, and the distribution of food retailers. First, our map displaying median household income demonstrates the income gap experienced in the Mission today. Median income in tracts ranged from 40,000-60,000 USD on the lower end, to 160,000-180,000 USD on the upper end, with five other ranges in between. The lowest and highest median income tracts share a corner in the Northwestern quadrant of the Mission, demonstrating the proximity of the income gap in the Mission District.

Our racial typography map revealed that Hispanic or Latinx and White are the majority races in the Mission District. There are ten tracts that were predominantly white and seven tracts that are predominantly Hispanic or Latinx. These findings demonstrate that a degree of racial diversity exists in the Mission District, albeit largely between two races. If you compare the Median Household Income map with the Racial Typography map it is apparent that there is some degree of correlation between lower-income census tracts and tracts that are predominately Hispanic/Latinx. This may be an important trend when researching levels of gentrification within the Mission District. However, while our findings display that there is a degree of racial integration within the Mission District, they do not reveal the degree of segregation occurring within each tact. Further research would be needed to understand the landscape of segregation within each tract. 

When we examined the socio-economic landscape of the Mission District in relation to the distribution of food retailers, we observed that there seems to be a level of correlation between the number of food retailers and tracts that are predominantly Hispanic and Latinx. If the maps we generated are compared to street maps (we direct readers to use Google Maps or other mapping software), the majority of food retailers are located along Mission Street and 24th Street--both streets are hubs for Latinx populations and culture. Our Racial Typography & Food Retailers map exemplifies that there may be a correlation between the location and quantity of food retailers in the Mission District and racial majority; however,  our Median Household Income & Food Retailers map doesn't demonstrate the same level of correlation. In the later case, the location of food retailers may have more to do with business corridors than median household income distribution.

In conclusion, we were able to replicate multiple maps used in researching the socio-economic landscapes of the Mission District. We were then able to start research, and open a discussion, about how this correlates with the distribution of food retailers within the district. Our work should be scene as a catalyst for future research on the food retail landscape of the Mission District. For example, it should be stated that in our maps all food retailers--regardless of size, stock, or purpose--are resembled by a single black dot. However, there is great diversity among food retailers. Some may be supermarkets, convenience stores, or corner markets. Additionally, some may cater to the general public, while others may sell specialty or ethnic foods. These kinds of differences would be important to know before drawing any conclusions regarding the distribution of food retailers in the Mission District in relation to socio-economic landscapes. Using our research as a launching pad, it is our hope that researches take our findings and code to further examine the role of food retailers in the socio-economic fabric of the Mission District.

#Research Data (DRAFT)

```{r}
retail_data <- read_csv("Food Retail Final List (Edit 3_30).csv")
retail_data

```
